New insights into the Weddell Sea ecosystem applying a network approach

### 2022-11-23 Manuscript/WordFormat.Rmd
Minor changes
2nd version of MS

The network approach has proved to shade light on the structure and stability of complex marine ecosystems. In recent years, such approach was succesfully applied to the study of polar ecosystems, improving our knowledge on how they might respond to ongoing environmental changes. The Weddell Sea, particularly the high continental shelf area (74-78ºS), is one of the most studied areas outside the Antarctic Peninsula in the Southern Ocean. However, few researches have considered the known complexity of the Weddell Sea. Understanding and projecting impacts of change in this ecosystem requires a framework considering the full complexity of it. Here we focus on the trophic interactions that underpin ecosystem structure and stability. We analysed the Weddell Sea food web, composed by 490 species and 16041 predator-prey interactions, which is the highest-resolved Antarctic food web constructed so far. The objective of our work was threefold: 1) estimate the strength for each interaction in the food web, 2) characterise the species considering weighted and unweighted properties, and 3) analyse the species' role in the stability of the food web. On the one hand, we found that the distribution of the interaction strength at the food web level is asymmetric, where weak interactions are prevalent. Including such information as a (weighted) property for species we detected a positive relationship between species mean interaction strength and two unweighted properties, trophic level and the total number of interactions. We also found that only few species are key in terms of food web stability, presenting particular weighted and unweighted properties. In the same analysis we have integrated food web and species information, enabling a more complete assessment of the ecosystem structure and function, likely highlighting the ecological processes at play in the Weddell Sea. We consider that our results provide new insights important for the development of effective policies and management strategies, particularly given the ongoing initiative to implement a Marine Protected Area (MPA) in the Weddell Sea.

@@

We analysed the food web of the Weddell Sea by: a) estimating the strength of each interaction; b) studying the species properties in a network framework; and c) comparing the stability of the food web after performing species extinction in silico simulations.

### Interaction strength estimation and distribution

To estimate the strength of each interaction in the food web we followed the methodology proposed by @Pawar2012. The minimum data requirements are: body mass of the consumer (predator) and resource (prey), and the interaction dimensionality classified as 2 or 3 dimensions. GATEWAy v.1.0 does provide the mean mass for consumers and resources (except for 'detritus' and 'sediment') and the dimensionality for the majority of the interactions, though the latter is missing in some cases (924 interactions). To complete such missing data, we used information about movement type for consumer and resource included in GATEWAy. Thus, we classified the interaction as 2D if both consumer and resource move in 2D (e.g., both are sessile or walking) or if a consumer moves in 3D and a resource in 2D (e.g., swimming consumer and sessile/walking resource). The interaction was classified as 3D if both consumer and resource move in 3D (e.g., both swimming) or if the consumer moves in 2D and the resource in 3D (e.g., sessile/walking consumer, swimming resource) [@Pawar2012].

@@

In order to individually characterize the species of the food web, we considered weighted and unweighted properties (Figure 2). The former is based on the estimation of the interaction strength described in the previous section. The latter is related to properties commonly used in qualitative or topological (presence/absence of interaction) food web studies [@Martinez1991; @Dunne2002; @Borrelli2014].

@@

As the weighted property we took into account the total mean interaction strength, meaning the average strength of all species' interactions for each species. On the other hand, we considered the following unweighted properties: a) degree or the total number of trophic interactions, summing up in- and out-coming interactions (role as predator and prey, respectively); b) trophic level or the position in the food web relative to primary producers/detritus; and c) trophic similarity or the trophic overlap based on shared and unique resources and consumers. The following are arguments to have selected the unweighted properties. The degree has often been equated with importance to the structure and functioning of a community, i.e. perturbations to high-degree species may therefore have larger effects on the food web than perturbations to low-degree species [@Dunne2002a; references in @Cirtwill2018a]. The trophic level offers information about how important a species is to its biotic community, i.e. top predators and primary producers are expected to have particularly large effects on the rest of their communities through top-down and bottom-up control, respectively [references in @Cirtwill2018a]. The trophic similarity is an index of trophic overlap considering the set of prey and predators for a pair of species; it measures one of the most important aspects of species’ niches, the trophic niche, and functional aspects of biodiversity [@Martinez1991; @Williams2000].

Furthermore, we took into account the species' habitat, which describes the physical position of a species within the ecosystem. Species were categorized as: 1) benthic, if the species lives on the seafloor; 2) pelagic, if the species lives close to the surface; 3) benthopelagic, if it moves between and connects the mentioned environments; 4) demersal, if it lives and feeds on or near the bottom of the sea; and 5) land-based, if the consumer is not aquatic but feeds predominantly in the marine realm. These data was taken from @Jacob2011.

With the aim of studying the relationship between the interaction strength of the species (weighted property) and its unweighted properties we performed linear regression analyses between the log(mean interaction strength) and each of the mentioned unweighted properties. We also explored such weighted property with species habitat.

@@

### Interaction strength

In this work we have estimated the interaction strength for the most highly-resolved marine food web to date, which comprises 490 species and 16041 predator-prey interactions. The distribution of the interaction strength best fit to a log-Normal model, meaning that there is a prevalent skew towards weaker interactions (Figure 3, Table 1).

```{r echo=FALSE, fig.align='center', fig.cap="Figure 3. Frequency distribution of interaction strengths for the Weddell Sea food web. Total number of interactions = 16041."}

@@

```{=tex}
\begin{table}[t]
\caption{Model comparison for the distribution of interaction strengths of the Weddell Sea food web. Order by best fit. References: df = degrees of freedom, AIC = Akaike Information Criterion, deltaAIC = difference with best fit. Log-Normal is the best model.}
\begin{tabular}{l c c c}
\tophline

\textbf{Model} & \textbf{df} & \textbf{AIC} & \textbf{deltaAIC} \\
\middlehline
log-Normal & 2 & -359738 & 0 \\
\middlehline
Gamma & 2 & -359714.1 & 23.89 \\
\middlehline
Power-law & 2 & -350667 & 9070.97 \\
\middlehline
Exponential & 1 & -327606.5 & 32131.54 \\
\middlehline
Normal & 2 & -291407.8 & 68330.21 \\
\middlehline
Uniform & 2 & -248179 & 111559.02 \\

\bottomhline
\end{tabular}
\end{table}
```

load("../Manuscript/ModelFit.rda")
knitr::kable(IS_fit %>% arrange(., deltaAIC), digits = c(0,0,2,2), col.names = c("Model", "df", "AIC", "deltaAIC"), caption = "Model comparison for the distribution of interaction strengths of the Weddell Sea food web. Order by best fit. References: df = degrees of freedom, AIC = Akaike Information Criterion, deltaAIC = difference with best fit. Log-Normal is the best model.")

Following the above results for the distribution of the interaction strengths, we found that the species mean interaction strength was different. We also found that such weighted property is related more or less with unweighted properties (Figure 4A-D). In this regard, interaction strength - trophic level showed a significant positive linear regression: the higher the trophic level of the species, the higher its mean interaction strength. We also found a significant but less evident positive relationship with degree or the total number of interactions. On the other hand, there was no significant relationship with trophic similarity. Considering species habitat, the "Benthopelagic" and "Pelagic" categories presented the two species with the highest mean interaction strength (Orcinus orca and Mesonychoteuthis hamiltoni, respectively), though "Demersal" and "Land-based" habitats concentrated the majority of the species with relatively higher interaction strength. Species inhabiting the benthic realm showed the lowest mean interaction strength (Figure 4D).

@@

It's important to note that since the proportion of Jacobians that were locally stable or QSS was zero for the Weddell Sea food web, we considered the mean maximum eigenvalue as the stability index. When performing the extinction simulations, our results showed that the majority of the species had no significant impact on the stability of the food web when extinct (Figure 5). This is shown in Figure 5, where most of the points lie around the zero value of the 'Stability difference' y axis, meaning that the stability for the food web minus a given species is similar (not significant) to the stability for the whole network. However, there were few species that when become extinct altered significantly the stability of the food web. Most of such species generated positive significant differences in the stability of the food web, which mean that they increased the network's stability when deleted. There were two exceptions, Pagetopsis macropterus (demersal fish) and Maxilliphimedia longipes (benthopelagic amphipod), that when deleted decreased the stability. Overall, 15 out of 490 species (3.06%) gave rise to significant changes in the food web's stability (Table 2).

```{r echo=FALSE, fig.align='center', fig.cap="Figure 5. Quasi-Sign Stability (QSS) difference between the whole Weddell Sea food web (n = 490) and the food web without one species (n = 489) for weighted (interaction strength) and unweighted species properties, and habitat. Point color indicates the impact on the QSS; if significant the extinction of that species altered the stability (QSS) of the food web."}

@@

After exploring the stability difference against the species properties (Figure 5), we found that those species that generated a significant impact on the stability of the food web were characterized by: 1) high mean interaction strength; 2) mid to high trophic levels (TL > 3.2); 3) relatively high number of interactions (Degree > 25); and 4) mid to low trophic similarity (TS < 0.16). Habitat wise, species with a significant impact on the stability were present in all habitats, except for the benthic realm. Table 2 shows these results for such species.

@@

Using the highest-resolved marine polar food web up to date, we show that 1) the distribution of the interaction strength at the food web level is asymmetric, where weak interactions are prevalent; that 2) there is a positive relationship between species mean interaction strength and two unweighted properties, trophic level and degree; and that 3) only few species are key in terms of food web stability, presenting particular weighted and unweighted properties. It's also noteworthy, that our study is the first of its type to consider the dimensionality of the interaction (2 or 3 dimensions) to estimate predator-prey relationships within a food web framework. After analysing more than 2900 predator-prey interactions from several ecosystems, @Pawar2012 concluded that interaction dimensionality is a critical factor driving consumer-resource dynamics, which will lead to better predictions of food web and ecosystem functioning. This arises as crucial to better understand how a complex system such as the Weddell Sea might respond to environmental change, which is an ongoing issue in that region of the Antarctic [@Gutt2022 and references herein].

Our results regarding the distribution of the interaction strengths are consistent with prior theoretical and empirical studies [e.g. @McCann1998; @Neutel2002; @Emmerson2004; @Wootton2005]. Within a system of a given complexity, commonly only a few strong species interactions are present with most interactions being weak. Historically, after @May1972 installed the debate complexity-stability, the results about the distribution of the interaction strength has been interpreted as an explanation for the persistence of complex food webs in nature [@Bascompte2005; @Allesina2015; @Nilsson2016]. Here we showed that this pattern is also prevalent in one of the most complex empirical marine food webs that was studied so far, with 490 species and 16041 predator-prey interactions.

Our approach to study species properties follows @Kortsch2021 recommendation that suggests to employ a range of descriptors from both unweighted and weighted food web in order to characterise the dynamic and multifaceted nature of structural and functional ecosystem changes. We went a step forward and analysed the relationship between weighted and unweighted descriptors. In this regard, our results show a positive relationship between interaction strength (weighted) and trophic level (unweighted), and between interaction strength and degree (total number of interactions). The former relationship might contradict those studies that suggest that mid-trophic level species are involved in the major pathways of energy flow in high-latitude marine ecosystems [@Pinkerton2014; @Murphy2016; @McCormack2020; @Riccialdelli2020]. This could be explained by the lack of species biomass data in our interaction strength estimations; the methodology we applied here [@Pawar2012] allows to include empirical data for the density of the resource, though not for all food web species. Unfortunately, the lack of individual data for the entire Weddell Sea food web hampers any alternative approach. On the other hand, the positive relationship between interaction strength and degree reinforces the importance of species with many interactions: species with high degree (hubs) have a large impact on overall food web structure and functioning [@Dunne2002a; @Kortsch2015]. In the Weddell Sea, species with high degree also tend to have high mean interaction strengths. This information on the quantity and quality of interactions and its relationship enables a robust assessment of the species' role in the stability of the food web [@Cirtwill2018a].

Our results show that only few species are key in terms of stability in the Weddell Sea food web. This is in concordance with studies of complex empirical food webs in marine ecosystems in the Arctic and other locations in Antarctica [@Kortsch2015; @Marina2018; @Rodriguez2022]. In the present work we identify that such key species present particular weighted and unweighted properties: 1) high mean interaction strength; 2) mid to high trophic level; 3) relatively high number of interactions; and 4) mid to low trophic similarity. The first attempt to gain insight into the role of species in the Weddell Sea applying a complex approach was that of @Jacob2011. These authors used information on body size, predatory and feeding type, and environment (habitat) to study species' roles. After performing sequential extinction simulations, @Jacob2011 found that larger bodied species can be lost without causing a direct collapse of the network topology. A major caveat of this finding, already recognized by the authors, is that population dynamics were ignored and therefore no top-down extinctions, or other indirect effects, could occur. In our study we did consider such effects by including information on the species interaction strength, which is of paramount importance when analysing the response of perturbations in ecological communities [@McCann1998; @Montoya2009; @Novak2011]. Thus, our study suggests that species with high mean interaction strength and high trophic level need to be considered with particular attention when trying to predict the effects of perturbations on the Weddell Sea ecosystem. Furthermore, this arises as a greater issue since we have identified that these species present a mid to low trophic similarity, meaning that few other species of the food web can occupy the same trophic role. In this sense, @Murphy2016's review emphasized that polar pelagic ecosystems are particularly sensitive to change due to a low functional redundancy at key trophic levels. It's important to note that we provide a broader analysis in this regard because we have included species from all habitats (pelagic, benthic, land-based).

@@

>>>>

Network approaches can shed light on the structure and stability of complex marine communities. In recent years, such approaches have been successfully applied to study polar ecosystems, improving our knowledge on how they might respond to ongoing environmental changes. The Weddell Sea is one of the most studied marine ecosystems outside the Antarctic Peninsula in the Southern Ocean. Yet, few studies consider the known complexity of the Weddell Sea food web which in its current form comprises 490 species and 16041 predator-prey interactions. Here we analysed the Weddell Sea food web, focusing on trophic interactions that underpin ecosystem structure and stability. We estimated the strength for each interaction in the food web, characterised species position in the food web using unweighted and weighted properties, and analysed species’ roles with respect to the stability of the food web. On one hand, we found that the distribution of the interaction strength at the food web level is asymmetric, where weak interactions are prevalent. Including such information as a (weighted) property for species we detected a positive relationship between species mean interaction strength and two unweighted properties, trophic level and the total number of interactions. We also found that only a few species are key in terms of food web stability, presenting high mean interaction strength, mid to high trophic level, relatively high number of interactions, and mid to low trophic similarity. In the same analysis we have integrated food web and species information, enabling a more complete assessment of the ecosystem structure and function, likely highlighting the ecological processes at play in the Weddell Sea. We consider that our results provide new insights important for the development of effective policies and management strategies, particularly given the ongoing initiative to implement a Marine Protected Area (MPA) in the Weddell Sea.

@@

#### Interaction strength estimation and distribution

To estimate the strength of each pairwise interaction in the food web we followed the methodology proposed by @Pawar2012. The minimum data requirements are body mass of the consumer (predator) and resource (prey), and the interaction dimensionality (ID) classified as 2 or 3 dimensions. The ID is defined as the dimension of the search space of the predator, that is equivalent to  the movement space of the prey. Thus, the ID is classified as 2D when both predator and prey move in 2D (e.g., both are benthic) or if a predator moves in 3D and a prey in 2D (e.g., pelagic predator on benthic prey). The ID is classified as 3D when both predator and prey move in 3D (e.g., both pelagic) or if the predator moves in 2D and the prey in 3D (e.g., benthic predator, pelagic prey) [@Pawar2012]. GATEWAy v.1.0 provides information on the mean body mass for consumers and resources, except for ‘detritus’ and ‘sediment’, and the dimensionality for the majority of the interactions, though the latter is missing in some cases (924 interactions). To complete the missing data on species ‘dimensionality’, we used information about the movement type of predators and prey  included in GATEWAy.

@@

To characterise the role of each species in the food web, we considered unweighted and weighted food web properties (Figure 2). Unweighted properties are related to properties commonly used in qualitative food web studies and only describe the presence or absence of interactions without any information on strength between a pairwise species link [@Martinez1991; @Dunne2002; @Borrelli2014]. In contrast, weighted properties capture the importance of interaction strength.

@@

To assess species roles as a function of the weighted food web, we focused on mean interaction strength defined as the average strength of all interactions for a given species. Further we calculated three unweighted species properties: a) species degree, i.e., the sum of in- and out-going interactions ; b) trophic level ; and c) trophic similarity, i.e., the trophic overlap based on shared and unique resources and consumers. These metrics were chosen to assess a species role based on the unweighted food web. The species degree has often been equated with species importance to the structure and functioning within a food web, i.e. perturbations to high-degree species may therefore have more significant effects on the food web robustness to perturbations than low-degree species [@Dunne2002a; references in @Cirtwill2018a]. The trophic level offers information about how important a species is to its biotic community, i.e., top predators and primary producers are expected to have particularly large effects on the rest of their communities through top-down and bottom-up control, respectively [references in @Cirtwill2018a]. Trophic similarity is an index of trophic overlap considering the set of prey and predators for a pair of species; it measures one of the most important aspects of species’ niches, the trophic niche, and functional aspects of biodiversity [@Martinez1991; @Williams2000].

Furthermore, we took a species’ habitat into account, which describes the physical position of a species within the ecosystem. Species were categorised as: 1) benthic, if a species lives on  the seafloor; 2) pelagic, if a species lives close to the surface; 3) benthopelagic, if it moves between and connects the mentioned environments; 4) demersal, if it lives and feeds on or near the bottom of the sea; and 5) land-based, if the consumer is not strictly aquatic but feeds predominantly on marine species. Species habitat affiliation was retrieved from @Jacob2011.

With the aim of studying the relationship between the interaction strength of the species (weighted property) and its unweighted properties we performed linear regression analyses between the log mean interaction strength and each of the mentioned unweighted properties. Thus we considered the interaction strength as the dependent variable and the given unweighted property as the independent variable, and obtained the coefficients (slope and intercept) for the linear model. Models were fitted using the least squares approach. We also explored the mean interaction strength distribution with the species habitat.

@@

### Interaction strength distribution

The statistical distribution that best fitted the empirical interaction strength distribution was a ‘gamma’ due to the high proportion of weak interactions and the existence of a few strong interactions (Figure 3, Table S3).

```{r echo=FALSE, fig.align='center', fig.cap="Figure 3. Frequency distribution of interaction strengths for the Weddell Sea food web. Total number of interactions = 16041. The distribution was best fitted to a ‘gamma’ model."}

@@

### Species’ role related to their mean interaction strength

We found that the species’ mean interaction strength (weighted property) shows different relationships with the unweighted properties analysed (Figure 4A-D). In this regard, there is a positive relationship between interaction strength and trophic level, i.e., the higher the trophic level of the species, the higher its mean interaction strength. We also found a significant but less evident positive relationship with species degree. Contrary, there was no significant relationship between mean interaction strength and trophic similarity. Considering species habitat affiliation, the “Benthopelagic” and “Pelagic” categories contained the two species with the highest mean interaction strength, the killer whale Orcinus orca and the colossal squid Mesonychoteuthis hamiltoni, respectively. However, the majority of the species with relatively higher interaction strength belonged to the “Demersal” and “Land-based” habitats groups. Species inhabiting the benthic realm showed the lowest mean interaction strength (Figure 4D).

@@

### Species impact on food web stability

Our extinction analyses showed that the majority of species had no significant impact on food web stability after being removed (Figure 5). Most of the species (black points in figure 5) did not change the stability of the network considerably after being removed, except for a few species. Only 15 out of 490 species (3.06%) gave rise to significant changes in the food web’s stability after their removal (Table 2). Most of these species had a positive impact on food web stability, i.e., network stability increased after their removal. Only two species significantly decreased network stability after being removed, the demersal fish Pagetopsis macropterus and the benthopelagic amphipod Maxilliphimedia longipes.

```{r echo=FALSE, fig.align='center', fig.cap="Figure 5. Stability  difference (mean maximum eingenvalue) between the whole Weddell Sea food web (n = 490) and the food web minus one species (n = 489) for weighted (interaction strength) and unweighted species properties, and habitat. Point color indicates the impact on the stability; if significant the extinction of that species altered the stability of the food web."}

@@

After exploring the stability difference against the species properties (Figure 5), we found that those species that generated a significant impact on the stability of the food web were characterised by: 1) high mean interaction strength; 2) mid to high trophic levels (TL > 3.2); 3) relatively high number of interactions (Degree > 25); and 4) mid to low trophic similarity (TS < 0.16). Habitat wise, species with a significant impact on the stability were present in all habitats, except for the benthic realm. Table 2 shows these results for such species.

@@

### Many weak and a few strong interactions

Our analyses show that the distribution of species interaction strength at the network level is asymmetric, i.e., the Weddell Sea food web contains many weak interactions and only a few strong ones. This finding is consistent with many previous theoretical and empirical studies [e.g. @McCann1998; @Neutel2002; @Emmerson2004; @Wootton2005; @Kortsch2021]. The asymmetric distribution of interaction strength in food webs has been interpreted as an explanation for the persistence of complex communities in nature [@Bascompte2005; @Allesina2015; @Nilsson2016]. Here we show that this pattern is also prevalent in one of the most complex empirical (marine) food webs to date, comprising 490 species and 16041 predator-prey interactions. This finding reinforces the call for the inclusion of interaction strength in food web studies to better understand the ecosystem functioning, and species and whole network responses to environmental perturbations.

### Species’s role related to their mean interaction strength

We employed a range of descriptors using both unweighted and weighted food web properties to characterise the dynamic and multifaceted nature of the Weddell Sea food web. Our results show a positive relationship between interaction strength and trophic level, and between interaction strength and species degree. Mean interaction strength increases with trophic level and species degree. The former relationship might contradict those studies that suggest that mid-trophic level species are involved in the major pathways of energy flow in high-latitude marine ecosystems [@Pinkerton2014; @Murphy2016; @McCormack2020; @Riccialdelli2020]. This could be explained by the lack of species biomass data in our interaction strength estimations; the methodology we applied here [@Pawar2012] allows empirical data for the density of the resource to be included, though this data is not for the majority of  food web species at the study site. On the other hand, the positive relationship between interaction strength and degree reinforces the importance of species with many interactions: species with high degree (hubs) have a large impact on overall food web structure and functioning [@Dunne2002a; @Kortsch2015]. In the Weddell Sea, species with high degree also tend to have high mean interaction strengths. This information on the quantity and quality of interactions and its relationship enables a robust assessment of the species’ role in the stability of the food web [@Cirtwill2018a].

### Species impact on food web stability

Only a few species play a key role with respect to the Weddell Sea food web stability, according to the mean maximum eingenvalue stability index employed in this study. This is in concordance with other studies on complex empirical food webs in marine ecosystems in the Arctic and other locations in Antarctica [@Kortsch2015; @Marina2018; @Rodriguez2022]. These key species are characterised by a particular set of food web properties: high to mean interaction strength; mid to high trophic level; relatively high number of interactions; and mid to low trophic similarity. In a previous study on sequential extinction simulations for the Weddell Sea food web, it was found  that larger bodied-sized species could be lost without causing a collapse of the network. A major caveat of this finding, also recognised by the authors, was that population dynamics were ignored and hence no top-down extinctions, or other indirect effects, could occur. In our study we considered such top-down effects by including information on the species interaction strength, which is of paramount importance when analysing the response of perturbations in ecological communities [@McCann1998; @Montoya2009; @Novak2011]. Thus, our study suggests that species with high mean interaction strength and high trophic level need to be considered with particular attention when trying to predict the effects of perturbations on the Weddell Sea ecosystem. This conclusion is further reinforced by the finding that these species have mid to low trophic similarity, which means that few other species of the food web can occupy the same trophic role. In a review, it was emphasised that polar pelagic communities are particularly sensitive to changes due to a low functional redundancy at key trophic levels [@Murphy2016]. Here we provide broader analyses of species impact on food web robustness by including species from all habitats (benthic, pelagic and land-based). This suggests that the sensitivity of marine polar ecosystems to environmental perturbations is a concern also beyond the pelagic realm.

@@

## Extras

It’s also noteworthy that our study is the first of its type to consider the dimensionality of the interaction (2 or 3 dimensions) to estimate predator-prey relationships within a food web framework. After analysing more than 2900 predator-prey interactions from several ecosystems, Pawar, Dell, and Van M. Savage (2012) concluded that interaction dimensionality is a critical factor driving consumer-resource dynamics, which will lead to better predictions of food web and ecosystem functioning. This arises as crucial to better understand how a complex system such as the Weddell Sea might respond to environmental change, which is an ongoing issue in that region of the Antarctic (Gutt et al. 2022 and references herein).
